---
output:
  html_document: default
  pdf_document: default
---

```{r}
cat("\014") # clear console
```

Lesson 3
========================================================
### Reading in Data
```{r}
getwd()
list.files()
pf <- read.csv('pseudo_facebook.tsv', sep = '\t')
names(pf)
```



***

### What to Do First?
Notes:

***

### Pseudo-Facebook User Data
Notes:

```{r Pseudo-Facebook User Data}

```

***

### Histogram of Users' Birthdays
Notes:


### Install Packages

#install.packages('ggplot2', dependencies = TRUE)
library(ggplot2)

#install.packages('ggthemes', dependencies = TRUE)
library(ggthemes)


```{r Histogram of Users\' Birthdays}
theme_set(theme_minimal(10))

names(pf)
qplot(x = dob_day, data = pf) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = 1:31)

ggplot(aes(x = dob_day), data = pf) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = 1:31)
```

***

#### What are some things that you notice about this histogram?
Response: Disproportionately large numbers on the 1 and low numbers on the 31st.  Not all months have 31 days.

***

### Moira's Investigation
Notes:

***

### Estimating Your Audience Size
Notes:

***

#### Think about a time when you posted a specific message or shared a photo on Facebook. What was it?
Response:

#### How many of your friends do you think saw that post?
Response:

#### Think about what percent of your friends on Facebook see any posts or comments that you make in a month. What percent do you think that is?
Response:

***

### Perceived Audience Size
Notes:

***
### Faceting
Notes:

```{r Faceting}
qplot(x = dob_day, data = pf) +
  scale_x_continuous(breaks = 1:31) +
  facet_wrap(~dob_month, ncol = 3)

ggplot(data = pf, aes(x = dob_day)) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = 1:31) +
  facet_wrap(~dob_month)

theme_set(theme_minimal(5))

```

#### Let’s take another look at our plot. What stands out to you here?
Response: January 1 is the >>> than other days

***

### Be Skeptical - Outliers and Anomalies
Notes:

***

### Moira's Outlier
Notes:
#### Which case do you think applies to Moira’s outlier?
Response:

***

### Friend Count
Notes:

#### What code would you enter to create a histogram of friend counts?

```{r Friend Count}
theme_set(theme_minimal(10))

qplot(x = friend_count, data = pf)

ggplot(aes(x = friend_count), data = pf) +
  geom_histogram()
```

#### How is this plot similar to Moira's first plot?
Response: High value at the beginning and a long tail

***

### Limiting the Axes
Notes:

```{r Limiting the Axes}
qplot(x = friend_count, data = pf, xlim = c(0, 1000))
```

```{r Limiting the Axes2}
qplot(x = friend_count, data = pf) +
  scale_x_continuous(limits = c(0, 1000))
```


```{r Limiting the Axes3}
ggplot(aes(x = friend_count), data = pf) +
  geom_histogram() +
  scale_x_continuous(limits = c(0, 1000))

```

### Exploring with Bin Width
Notes:

***

### Adjusting the Bin Width
Notes:

### Faceting Friend Count
```{r Faceting Friend Count}
# What code would you add to create a facet the histogram by gender?
# Add it to the code below.
qplot(x = friend_count, data = pf, binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50))
```

***
```{r Faceting Friend Count2}
# What code would you add to create a facet the histogram by gender?
# Add it to the code below.
qplot(x = friend_count, data = pf, binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender)
```



### Omitting NA Values
Notes:

```{r Omitting NA Values}

# What code would you add to create a facet the histogram by gender?
# Add it to the code below.
qplot(x = friend_count, data = subset(pf, !is.na(gender)), binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender)
```


```{r Omitting NA Values2}

# What code would you add to create a facet the histogram by gender?
# Add it to the code below.  Removes na values from all groups
qplot(x = friend_count, data = na.omit(pf), binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender)
```


***

### Statistics 'by' Gender
Notes:

```{r Statistics \'by\' Gender}
table(pf$gender)
by(pf$friend_count, pf$gender, summary)
```

#### Who on average has more friends: men or women?
Response: Women

#### What's the difference between the median friend count for women and men?
Response: 22

#### Why would the median be a better measure than the mean?
Response: Because high values, long tails, skew the mean higher.

***

### Tenure
Notes:

```{r Tenure}
qplot(x = tenure, data = pf, binwidth = 30, color = I('black'), fill = I('#BA55D3'))


ggplot(aes(x = tenure), data = pf) +
  geom_histogram(binwidth = 30, color = 'black', fill = '#099DD9')
```

***

#### How would you create a histogram of tenure by year?

```{r Tenure Histogram by Year}

```

***

### Labeling Plots
Notes:

```{r Labeling Plots}
qplot(x = tenure/365, data = pf,
      xlab = 'Number of years using FB',
      ylab = 'Number of users in sample',
      color = I('black'), fill = I('#F79420')) +
  scale_x_continuous(breaks = seq(1, 7, 1), lim = c(0, 7))


ggplot(aes(x = tenure / 365), data = pf) +
  geom_histogram(color = 'black', fill = '#BA55D3') +
  scale_x_continuous(breaks = seq(1, 7, 1), limits = c(0, 7)) +
  xlab('Number of years using Facebook') +
  ylab('Number of users in sample')

```

***

### User Ages
Notes:

```{r User Ages}
summary(pf$age)

qplot(x = age, data = pf, bins = 115,
      ylab = 'Count',
      xlab = 'Age',
      color = I('black'), fill = I('#BA55D3')) +
  scale_x_continuous(breaks = seq(0, 115, 5), lim = c(0, 115))


ggplot(aes(x = age), data = pf) +
  geom_histogram(binwidth = 1, fill = '#5760AB') +
  scale_x_continuous(breaks = seq(0, 115, 5))
```

#### What do you notice?
Response:

***

### The Spread of Memes
Notes:

***

### Lada's Money Bag Meme
Notes:

***

### Transforming Data
Notes:



### Install Packages

```{r}
#install.packages('gridExtra')
library(gridExtra)
```



```{r Transforming Data}
qplot(x = friend_count, data = pf)

summary(pf$friend_count)

summary(log10(pf$friend_count + 1))

summary(sqrt(pf$friend_count))
```


```{r Transforming Data3}
p1 = ggplot(aes(x = friend_count), data = pf) +
  geom_histogram(color = 'black', fill = '#BA55D3')

p2 = p1 + scale_x_log10()

p3 = p1 + scale_x_sqrt()

  
grid.arrange(p1, p2, p3, ncol = 1)
```


```{r Transforming Data4}
p1 = qplot(x = friend_count, data = pf)
p2 = qplot(x = log10(friend_count + 1), data = pf)
p3 = qplot(x = sqrt(friend_count), data = pf)

grid.arrange(p1, p2, p3, ncol = 1)

```
***

### Add a Scaling Layer
Notes:

```{r Add a Scaling Layer}
logScale = qplot(x = log10(friend_count), data = pf)

countScale = ggplot(aes(x = friend_count), data = pf) +
  geom_histogram() +
  scale_x_log10()

grid.arrange(logScale, countScale, ncol = 2)

qplot(x = friend_count, data = pf) +
  scale_x_log10()


ggplot(aes(x = friend_count), data = pf) +
  geom_histogram(color = 'black', fill = '#BA55D3') +
  scale_x_log10()
```

***


### Frequency Polygons

```{r Frequency Polygons}
qplot(x = friend_count, data = subset(pf, !is.na(gender)),
      binwidth = 10) +
  scale_x_continuous(lim = c(0, 1000), breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender)


qplot(x = friend_count, y = ..count../sum(..count..),
      data = subset(pf, !is.na(gender)),
      xlab = 'Friend Count',
      ylab = 'Proportion of Users with that friend count',
      binwidth = 10, geom = 'freqpoly', color = gender) +
  scale_x_continuous(lim = c(0, 1000), breaks = seq(0, 1000, 50))
```

***

### Likes on the Web
Notes:
```{r Likes on the Web}
qplot(x = www_likes, data = subset(pf, !is.na(gender)),
      binwidth = 10) +
  scale_x_continuous(lim = c(0, 100), breaks = seq(0, 100, 5)) +
  scale_y_continuous(lim = c(0, 4000)) +
  facet_wrap(~gender)

qplot(x = www_likes, y = ..count../sum(..count..),
      data = subset(pf, !is.na(gender)),
      xlab = 'www_likes',
      ylab = 'Proportion of Users with that www_likes count',
      geom = 'freqpoly', color = gender) +
  scale_x_continuous(lim = c(0, 100), breaks = seq(0, 100, 5))

qplot(x = www_likes,
      data = subset(pf, !is.na(gender)),
      xlab = 'www_likes',
      ylab = 'Proportion of Users with that www_likes count',
      geom = 'freqpoly', color = gender) +
  scale_x_log10()

qplot(x = www_likes, y = ..count../sum(..count..),
      data = subset(pf, !is.na(gender)),
      xlab = 'www_likes',
      ylab = 'Proportion of Users with that www_likes count',
      geom = 'freqpoly', color = gender) +
  scale_x_log10()

by(pf$www_likes, pf$gender, sum)

summary(pf$www_likes)
```


***

### Box Plots
Notes:

```{r Box Plots}
qplot(x = gender, y = friend_count,
      data = subset(pf, !is.na(gender)),
      geom = 'boxplot')
```

#### Adjust the code to focus on users who have friend counts between 0 and 1000.

```{r}
qplot(x = gender, y = friend_count,
      data = subset(pf, !is.na(gender)),
      geom = 'boxplot') +
  scale_y_continuous(lim = c(0, 1000))

qplot(x = gender, y = friend_count,
      data = subset(pf, !is.na(gender)),
      geom = 'boxplot', ylim = c(0, 1000))

#previous two methods remove data points from calculations

qplot(x = gender, y = friend_count,
      data = subset(pf, !is.na(gender)),
      geom = 'boxplot') +
  coord_cartesian(ylim = c(0, 1000))
```

***

### Box Plots, Quartiles, and Friendships
Notes:

```{r Box Plots, Quartiles, and Friendships}
qplot(x = gender, y = friend_count,
      data = subset(pf, !is.na(gender)),
      geom = 'boxplot') +
  coord_cartesian(ylim = c(0, 250))

by(pf$friend_count, pf$gender, summary)

```

#### On average, who initiated more friendships in our sample: men or women?
Response:
#### Write about some ways that you can verify your answer.
Response:
```{r Friend Requests by Gender}
qplot(x = gender, y = friendships_initiated,
      data = subset(pf, !is.na(gender)),
      geom = 'boxplot') +
  coord_cartesian(ylim = c(0, 150))

by(pf$friendships_initiated, pf$gender, summary)
```

Response:

***

### Getting Logical
Notes:

```{r Getting Logical}
summary(pf$mobile_likes)

summary(pf$mobile_likes > 0)

pf$mobile_check_in <- NA
pf$mobile_check_in <- ifelse(pf$mobile_likes > 0, 1, 0)
pf$mobile_check_in <- factor(pf$mobile_check_in)
summary(pf$mobile_check_in)

tot_mob_check_in = length(pf$mobile_check_in)
tot_mob_check_in

(sum(pf$mobile_check_in == 1)/tot_mob_check_in)*100
```

Response:

***

### Analyzing One Variable
Reflection:

***

rmarkdown::render('E:/Users/Trenton J. McKinney/PycharmProjects/UDACITY/01_Data_Analyst/05_R_Data_Analysis/eda-course-materials/lesson3/lesson3_student.rmd')